\textbf{Machine learning classification by fitting amplicon sequences to existing OTUs}

### 2023-07-05 submission/second_revision/manuscript.tex
fix citation error

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Machine learning classification by fitting amplicon sequences to existing OTUs},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

To test how the model performance compared between using \emph{de novo}
and reference-based clustering approaches, we used a publicly available
dataset of 16S rRNA gene sequences from stool samples of healthy
subjects (n = 226) as well as subjects with screen-relevant neoplasia
(SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{1}). For the \emph{de novo}
workflows, the 16S rRNA sequence data from all samples were clustered
into OTUs using the OptiClust algorithm in mothur 1 and the VSEARCH
algorithm used in QIIME2 (\protect\hyperlink{ref-rognes2016}{6},
\protect\hyperlink{ref-bolyen2019}{7}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). For traditional reference-based
clustering (database-reference-based), we used OptiFit to fit the
sequence data into OTUs based on the commonly used greengenes reference
database. To compare with another commonly used method, we also used
VSEARCH to map sequences to reference OTUs from the greengenes database
with the parameters used by QIIME2. Again, the data was then split into
training and testing sets, hyperparameters tuned, and performance
evaluated on the testing set (Figure 1B). In the OptiFit self-reference
workflow (self-reference-based), the data was split into a training and
a testing set. The training set was clustered into OTUs and used to
train a classification model. The OptiFit algorithm was used to fit
sequence data of samples not part of the training data into the training
OTUs and classified using the best hyperparameters (Figure 1C). For each
of the workflows the process was repeated for 100 random splits of the
data to account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is an
objective metric used to measure OTU cluster quality based on the
similarity of all pairs of sequences and whether they are appropriately
clustered or not (\protect\hyperlink{ref-westcott2015}{8}). We expected
the MCC scores produced by the OptiFit workflow to be similar to that of
\emph{de novo} clustering using the OptiClust algorithm. In the OptiFit
workflow the test data was fit to the clustered training data for each
of the 100 data splits resulting in an MCC score for each split of the
data. In the remaining workflows, the data was only clustered once and
then split into the training and testing sets resulting in a single MCC
score for each method. Indeed, the MCC scores were similar between the
OptiClust \emph{de novo} (MCC = 0.884) and OptiFit self-reference
workflows (average MCC = 0.879, standard deviation = 0.002). Consistent
with prior findings, the reference-based methods produced lower MCC
scores (OptiFit greengenes MCC = 0.786; VSEARCH greengenes MCC = 0.531)
than the \emph{de novo} methods (OptiClust \emph{de novo} MCC = 0.884;
VSEARCH \emph{de novo} MCC = 0.641)
(\protect\hyperlink{ref-sovacool2022}{5}). Another metric we examined
for the OptiFit workflow was the fraction of sequences from the test set
that mapped to the reference OTUs. Since sequences that did not map to
reference OTUs were eliminated, if a high percentage of reads did not
map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.7\%) of sequences in the
subsampled test set mapped to the reference OTUs. This number is higher
than the average fraction of reads mapped in the OptiFit greengenes
workflow (mean = 96.8\%, standard deviation = 3.5). These results
indicate that the OptiFit self-reference method performed as well as the
OptiClust \emph{de novo} method and better than using an external
database.

@@

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{9}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{10}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{11}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{5}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    The preprocessed data was fit to the clustered reference data using
    OptiFit with the method open to allow any sequences that did not map
    to the existing reference clusters would form new OTUs. The data was
    then subsampled to 10,000 reads and any novel OTUs from the test set
    were removed. The dataset was then split into two sets where 80\% of
    the samples were assigned to the training set and 20\% to the
    testing set. This process was repeated for each of the 100 splits
    resulting in 100 training and testing datasets.
  \end{enumerate}
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{7}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust 1 to generate OTUs. The data was subsampled to
  10,000 reads per sample and the resulting abundance tables were split
  into the training and testing sets. The process was repeated for each
  of the 100 splits resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{Machine Learning.} A random forest model was trained with the R
package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P-values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{Code Availability.} The analysis workflow was implemented in
Snakemake (\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis
were written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{9}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{6}), RStudio
(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023}

@@

\leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
B}, \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: a versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.}

\leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
MR}, \textbf{Bokulich NA}, \textbf{Abnet CC}, \textbf{Al-Ghalith GA},
\textbf{Alexander H}, \textbf{Alm EJ}, \textbf{Arumugam M},
\textbf{Asnicar F}, \textbf{Bai Y}, \textbf{Bisanz JE},
\textbf{Bittinger K}, \textbf{Brejnrod A}, \textbf{Brislawn CJ},
\textbf{Brown CT}, \textbf{Callahan BJ}, \textbf{Caraballo-Rodríguez
AM}, \textbf{Chase J}, \textbf{Cope EK}, \textbf{Da Silva R},
\textbf{Diener C}, \textbf{Dorrestein PC}, \textbf{Douglas GM},
\textbf{Durall DM}, \textbf{Duvallet C}, \textbf{Edwardson CF},
\textbf{Ernst M}, \textbf{Estaki M}, \textbf{Fouquier J},
\textbf{Gauglitz JM}, \textbf{Gibbons SM}, \textbf{Gibson DL},
\textbf{Gonzalez A}, \textbf{Gorlick K}, \textbf{Guo J},
\textbf{Hillmann B}, \textbf{Holmes S}, \textbf{Holste H},
\textbf{Huttenhower C}, \textbf{Huttley GA}, \textbf{Janssen S},
\textbf{Jarmusch AK}, \textbf{Jiang L}, \textbf{Kaehler BD},
\textbf{Kang KB}, \textbf{Keefe CR}, \textbf{Keim P}, \textbf{Kelley
ST}, \textbf{Knights D}, \textbf{Koester I}, \textbf{Kosciolek T},
\textbf{Kreps J}, \textbf{Langille MGI}, \textbf{Lee J}, \textbf{Ley R},
\textbf{Liu Y-X}, \textbf{Loftfield E}, \textbf{Lozupone C},
\textbf{Maher M}, \textbf{Marotz C}, \textbf{Martin BD},
\textbf{McDonald D}, \textbf{McIver LJ}, \textbf{Melnik AV},
\textbf{Metcalf JL}, \textbf{Morgan SC}, \textbf{Morton JT},
\textbf{Naimey AT}, \textbf{Navas-Molina JA}, \textbf{Nothias LF},
\textbf{Orchanian SB}, \textbf{Pearson T}, \textbf{Peoples SL},
\textbf{Petras D}, \textbf{Preuss ML}, \textbf{Pruesse E},
\textbf{Rasmussen LB}, \textbf{Rivers A}, \textbf{Robeson MS},
\textbf{Rosenthal P}, \textbf{Segata N}, \textbf{Shaffer M},
\textbf{Shiffer A}, \textbf{Sinha R}, \textbf{Song SJ}, \textbf{Spear
JR}, \textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Torres PJ},
\textbf{Trinh P}, \textbf{Tripathi A}, \textbf{Turnbaugh PJ},
\textbf{Ul-Hasan S}, \textbf{Hooft JJJ van der}, \textbf{Vargas F},
\textbf{Vázquez-Baeza Y}, \textbf{Vogtmann E}, \textbf{Hippel M von},
\textbf{Walters W}, \textbf{Wan Y}, \textbf{Wang M}, \textbf{Warren J},
\textbf{Weber KC}, \textbf{Williamson CHD}, \textbf{Willis AD},
\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhang Y}, \textbf{Zhu Q},
\textbf{Knight R}, \textbf{Caporaso JG}. 2019. Reproducible,
interactive, scalable and extensible microbiome data science using QIIME
2. Nature Biotechnology \textbf{37}:852--857.
doi:\href{https://doi.org/10.1038/s41587-019-0209-9}{10.1038/s41587-019-0209-9}.}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Machine learning classification by fitting amplicon sequences to existing OTUs},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

To test how the model performance compared between using \emph{de novo}
and reference-based clustering approaches, we used a publicly available
dataset of 16S rRNA gene sequences from stool samples of healthy
subjects (n = 226) as well as subjects with screen-relevant neoplasia
(SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{1}). For the \emph{de novo}
workflows, the 16S rRNA sequence data from all samples were clustered
into OTUs using the OptiClust algorithm in mothur
(\protect\hyperlink{ref-westcott2017}{6}) and the VSEARCH algorithm used
in QIIME2 (\protect\hyperlink{ref-rognes2016}{7},
\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). For traditional reference-based
clustering (database-reference-based), we used OptiFit to fit the
sequence data into OTUs based on the commonly used greengenes reference
database. To compare with another commonly used method, we also used
VSEARCH to map sequences to reference OTUs from the greengenes database
with the parameters used by QIIME2. Again, the data was then split into
training and testing sets, hyperparameters tuned, and performance
evaluated on the testing set (Figure 1B). In the OptiFit self-reference
workflow (self-reference-based), the data was split into a training and
a testing set. The training set was clustered into OTUs and used to
train a classification model. The OptiFit algorithm was used to fit
sequence data of samples not part of the training data into the training
OTUs and classified using the best hyperparameters (Figure 1C). For each
of the workflows the process was repeated for 100 random splits of the
data to account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is an
objective metric used to measure OTU cluster quality based on the
similarity of all pairs of sequences and whether they are appropriately
clustered or not (\protect\hyperlink{ref-westcott2015}{9}). We expected
the MCC scores produced by the OptiFit workflow to be similar to that of
\emph{de novo} clustering using the OptiClust algorithm. In the OptiFit
workflow the test data was fit to the clustered training data for each
of the 100 data splits resulting in an MCC score for each split of the
data. In the remaining workflows, the data was only clustered once and
then split into the training and testing sets resulting in a single MCC
score for each method. Indeed, the MCC scores were similar between the
OptiClust \emph{de novo} (MCC = 0.884) and OptiFit self-reference
workflows (average MCC = 0.879, standard deviation = 0.002). Consistent
with prior findings, the reference-based methods produced lower MCC
scores (OptiFit greengenes MCC = 0.786; VSEARCH greengenes MCC = 0.531)
than the \emph{de novo} methods (OptiClust \emph{de novo} MCC = 0.884;
VSEARCH \emph{de novo} MCC = 0.641)
(\protect\hyperlink{ref-sovacool2022}{5}). Another metric we examined
for the OptiFit workflow was the fraction of sequences from the test set
that mapped to the reference OTUs. Since sequences that did not map to
reference OTUs were eliminated, if a high percentage of reads did not
map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.7\%) of sequences in the
subsampled test set mapped to the reference OTUs. This number is higher
than the average fraction of reads mapped in the OptiFit greengenes
workflow (mean = 96.8\%, standard deviation = 3.5). These results
indicate that the OptiFit self-reference method performed as well as the
OptiClust \emph{de novo} method and better than using an external
database.

@@

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{12}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{5}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{13}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust
  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
  fit to the clustered reference data using OptiFit with the method open
  to allow any sequences that did not map to the existing reference
  clusters would form new OTUs. The data was then subsampled to 10,000
  reads and any novel OTUs from the test set were removed. The dataset
  was then split into two sets where 80\% of the samples were assigned
  to the training set and 20\% to the testing set. This process was
  repeated for each of the 100 splits resulting in 100 training and
  testing datasets.
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The dataset was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
  generate OTUs. The data was subsampled to 10,000 reads per sample and
  the resulting abundance tables were split into the training and
  testing sets. The process was repeated for each of the 100 splits
  resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{Machine Learning.} A random forest model was trained with the R
package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{14}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P-values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{15}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{Code Availability.} The analysis workflow was implemented in
Snakemake (\protect\hyperlink{ref-koster2012}{16}). Scripts for analysis
were written in R (\protect\hyperlink{ref-R2020}{17}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{18}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{19}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{20}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{21}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{22}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{23}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023}

@@

\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
OptiClust, an improved method for assigning amplicon-based sequence data
to operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}

\leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
B}, \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: a versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.}

\leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
MR}, \textbf{Bokulich NA}, \textbf{Abnet CC}, \textbf{Al-Ghalith GA},
\textbf{Alexander H}, \textbf{Alm EJ}, \textbf{Arumugam M},
\textbf{Asnicar F}, \textbf{Bai Y}, \textbf{Bisanz JE},
\textbf{Bittinger K}, \textbf{Brejnrod A}, \textbf{Brislawn CJ},
\textbf{Brown CT}, \textbf{Callahan BJ}, \textbf{Caraballo-Rodríguez
AM}, \textbf{Chase J}, \textbf{Cope EK}, \textbf{Da Silva R},
\textbf{Diener C}, \textbf{Dorrestein PC}, \textbf{Douglas GM},
\textbf{Durall DM}, \textbf{Duvallet C}, \textbf{Edwardson CF},
\textbf{Ernst M}, \textbf{Estaki M}, \textbf{Fouquier J},
\textbf{Gauglitz JM}, \textbf{Gibbons SM}, \textbf{Gibson DL},
\textbf{Gonzalez A}, \textbf{Gorlick K}, \textbf{Guo J},
\textbf{Hillmann B}, \textbf{Holmes S}, \textbf{Holste H},
\textbf{Huttenhower C}, \textbf{Huttley GA}, \textbf{Janssen S},
\textbf{Jarmusch AK}, \textbf{Jiang L}, \textbf{Kaehler BD},
\textbf{Kang KB}, \textbf{Keefe CR}, \textbf{Keim P}, \textbf{Kelley
ST}, \textbf{Knights D}, \textbf{Koester I}, \textbf{Kosciolek T},
\textbf{Kreps J}, \textbf{Langille MGI}, \textbf{Lee J}, \textbf{Ley R},
\textbf{Liu Y-X}, \textbf{Loftfield E}, \textbf{Lozupone C},
\textbf{Maher M}, \textbf{Marotz C}, \textbf{Martin BD},
\textbf{McDonald D}, \textbf{McIver LJ}, \textbf{Melnik AV},
\textbf{Metcalf JL}, \textbf{Morgan SC}, \textbf{Morton JT},
\textbf{Naimey AT}, \textbf{Navas-Molina JA}, \textbf{Nothias LF},
\textbf{Orchanian SB}, \textbf{Pearson T}, \textbf{Peoples SL},
\textbf{Petras D}, \textbf{Preuss ML}, \textbf{Pruesse E},
\textbf{Rasmussen LB}, \textbf{Rivers A}, \textbf{Robeson MS},
\textbf{Rosenthal P}, \textbf{Segata N}, \textbf{Shaffer M},
\textbf{Shiffer A}, \textbf{Sinha R}, \textbf{Song SJ}, \textbf{Spear
JR}, \textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Torres PJ},
\textbf{Trinh P}, \textbf{Tripathi A}, \textbf{Turnbaugh PJ},
\textbf{Ul-Hasan S}, \textbf{Hooft JJJ van der}, \textbf{Vargas F},
\textbf{Vázquez-Baeza Y}, \textbf{Vogtmann E}, \textbf{Hippel M von},
\textbf{Walters W}, \textbf{Wan Y}, \textbf{Wang M}, \textbf{Warren J},
\textbf{Weber KC}, \textbf{Williamson CHD}, \textbf{Willis AD},
\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhang Y}, \textbf{Zhu Q},
\textbf{Knight R}, \textbf{Caporaso JG}. 2019. Reproducible,
interactive, scalable and extensible microbiome data science using QIIME
2. Nature Biotechnology \textbf{37}:852--857.
doi:\href{https://doi.org/10.1038/s41587-019-0209-9}{10.1038/s41587-019-0209-9}.}
